**This is only a basic Mini-Project Template.**
**Delete this note and customize formatting according to your preferences and**
**the Mini-Project instructions. Your document must start the `---` below**
---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "KevMarcel0"
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

Remember that code blocks look like this: 



```{r}
library(tidycensus)
library(gt)

ny_income <- get_acs(
    c(median_household_income="B19013_001"), 
    state="NY", 
    geography="county", 
    year=2018, 
    survey="acs1"
)
```

```{r}
library(tidyverse)
library(tidycensus)
library(sf)
load_state_population <- function(year = 2015:2024) {
    
    # Create folder if it does not exist
    if(!dir.exists(file.path("data", "mp03"))){
        dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
    }
    
    # Download and combine data across years
    pop_data <- map_dfr(
        year,
        \(yr) {
            get_acs(
                geography = "state",
                variables = c(total_population = "B01003_001"),
                year = yr,
                geometry = TRUE,
                cache_table = TRUE
            ) |>
                transmute(
                    total_population = estimate,
                    state = NAME,
                    year = yr
                )
        }
    )
    
    pop_data
}
# Run function
state_pop <- load_state_population()
# Check result
glimpse(state_pop)
head(state_pop)

```



```{r}
#Task 2
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(openxlsx)

state_to_state_migration_2024 <- function() {
  
  if (!dir.exists(file.path("data", "mp03"))) {
    dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
  }
  
  xlsx_file <- file.path(
    "data",
    "mp03",
    "State_to_State_Migration_Table_2024_T13.xlsx"
  )
  
  if (!file.exists(xlsx_file)) {
    download.file(
      url = "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx",
      destfile = xlsx_file,
      mode = "wb",
      quiet = TRUE
    )
  }
  
  table_csv <- file.path("data", "mp03", "state_flows_2024_table.csv")
  supp_csv  <- file.path("data", "mp03", "state_flows_2024_supp.csv")
  
  if (!file.exists(table_csv) || !file.exists(supp_csv)) {
    
    table_df <- read.xlsx(
      xlsx_file,
      sheet = "Table",
      startRow = 8,
      colNames = TRUE
    ) |> as_tibble()
    
    supp_df <- read.xlsx(
      xlsx_file,
      sheet = "Supplemental - Current Res",
      startRow = 8,
      colNames = TRUE
    ) |> as_tibble()
    
    write.csv(table_df, table_csv, row.names = FALSE)
    write.csv(supp_df, supp_csv, row.names = FALSE)
  }
  
  states_with_dc <- c(
    state.name[1:8],
    "District of Columbia",
    state.name[9:50]
  )
  
  states_abbrev_with_dc <- c(
    state.abb[1:8],
    "DC",
    state.abb[9:50]
  )
  
  flow_data <- suppressMessages(
    read_csv(table_csv, show_col_types = FALSE)
  ) |>
    rename(
      state_current = X1,
      state_1y = X2,
      population = Estimate
    ) |>
    mutate(
      state_current = str_squish(as.character(state_current)),
      state_1y = str_squish(as.character(state_1y)),
      population = as.character(population)
    ) |>
    filter(
      !is.na(state_current),
      !is.na(state_1y),
      state_current %in% states_with_dc,
      state_1y %in% states_with_dc,
      population != "X"
    ) |>
    mutate(
      state_current = recode(state_current, !!!setNames(states_abbrev_with_dc, states_with_dc)),
      state_1y = recode(state_1y, !!!setNames(states_abbrev_with_dc, states_with_dc)),
      population = if_else(population == "N", "0", population),
      population = as.integer(population),
      year_current = 2024L,
      year_1y = 2023L
    ) |>
    filter(
      state_current != "DC",
      state_1y != "DC"
    ) |>
    select(state_current, state_1y, population, year_current, year_1y)
  
  stayed_data <- suppressMessages(
    read_csv(supp_csv, show_col_types = FALSE)
  )
  
  # adjust these names if needed after inspecting the supplemental CSV
  stayed_data <- stayed_data |>
    select(1:3) |>
    rename(
      state_current = 1,
      stayed_same_house = 2,
      moved_within_state = 3
    ) |>
    mutate(
      state_current = str_squish(as.character(state_current))
    ) |>
    filter(
      !is.na(state_current),
      state_current %in% states_abbrev_with_dc
    ) |>
    mutate(
      stayed_same_house = as.integer(replace_na(as.character(stayed_same_house), "0")),
      moved_within_state = as.integer(replace_na(as.character(moved_within_state), "0")),
      population = stayed_same_house + moved_within_state,
      state_1y = state_current,
      year_current = 2024L,
      year_1y = 2023L
    ) |>
    select(state_current, state_1y, population, year_current, year_1y)
  
  bind_rows(flow_data, stayed_data)
}
```

```{r}
flows_2024 <- state_to_state_migration_2024()

dim(flows_2024)
head(flows_2024)
```


```{r}
# Task 3
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(openxlsx)
library(vctrs)

state_to_state_migration_2023 <- function() {
  
  if (!dir.exists(file.path("data", "mp03"))) {
    dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
  }
  
  xlsx_file <- file.path(
    "data",
    "mp03",
    "State_to_State_Migration_Table_2023_T13.xlsx"
  )
  
  if (!file.exists(xlsx_file)) {
    download.file(
      url = "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2023/state-to-state-migration/State_to_State_Migration_Table_2023_T13.xlsx",
      destfile = xlsx_file,
      mode = "wb",
      quiet = TRUE
    )
  }
  
  states_with_dc <- c(
    state.name[1:8],
    "District of Columbia",
    state.name[9:50]
  )
  
  states_abbrev_with_dc <- c(
    state.abb[1:8],
    "DC",
    state.abb[9:50]
  )
  
  name_to_abbrev <- setNames(states_abbrev_with_dc, states_with_dc)
  
  # ---------------------------
  # 1. Stayed within the state
  # First 6 columns only
  # ---------------------------
  stay_data <- read.xlsx(
  xlsx_file,
  rows = 9:65,
  cols = 1:6,
  colNames = FALSE
) |>
  as_tibble(.name_repair = "minimal") |>
  select(
    state_current = 1,
    stayed_same_house = 2,
    moved_within_state = 3
  ) |>
  mutate(
    state_current = stringr::str_squish(as.character(state_current))
  ) |> 
  filter(!is.na(state_current)) |>
  filter(state_current %in% states_with_dc) |>
  mutate(
    state_current = recode(state_current, !!!name_to_abbrev),
    stayed_same_house = as.integer(replace_na(as.character(stayed_same_house), "0")),
    moved_within_state = as.integer(replace_na(as.character(moved_within_state), "0")),
    population = stayed_same_house + moved_within_state,
    state_1y = state_current,
    year_current = 2023L,
    year_1y = 2022L
  ) |>
  filter(state_current != "DC") |>
  select(state_current, state_1y, population, year_current, year_1y)
  
   # ---------------------------
  # 2. Actual state-to-state flows
  # Columns J:DG only
  # ---------------------------
  row_states <- read.xlsx(
    xlsx_file,
    rows = 9:65,
    cols = 1,
    colNames = FALSE
  ) |>
    as_tibble(.name_repair = "minimal") |>
    rename(state_1y = 1)
  
  flow_wide <- read.xlsx(
    xlsx_file,
    rows = 9:65,
    cols = 10:111,
    colNames = FALSE
  ) |>
    as_tibble(.name_repair = "minimal")
  
  # make sure both have the same number of rows
  n_rows <- min(nrow(row_states), nrow(flow_wide))
  row_states <- slice(row_states, 1:n_rows)
  flow_wide  <- slice(flow_wide, 1:n_rows)
  
  colnames(flow_wide) <- vctrs::vec_interleave(states_with_dc, NA_character_)
  
 flow_data <- bind_cols(row_states, flow_wide) |>
  select(state_1y, any_of(states_with_dc)) |>
  filter(!is.na(state_1y), state_1y %in% states_with_dc) |>
  mutate(
    across(any_of(states_with_dc), as.character)
  ) |>
  pivot_longer(
    cols = -state_1y,
    names_to = "state_current",
    values_to = "population"
  ) |>
  filter(
    !is.na(population),
    state_current != "District of Columbia",
    state_1y != "District of Columbia",
    state_current != state_1y
  ) |>
  mutate(
    state_current = recode(state_current, !!!name_to_abbrev),
    state_1y = recode(state_1y, !!!name_to_abbrev),
    population = if_else(population == "N", "0", population),
    population = as.integer(population),
    year_current = 2023L,
    year_1y = 2022L
  ) |>
  select(state_current, state_1y, population, year_current, year_1y)
}
```



```{r}
library(tidyverse)
state_to_state_migration <- function() {
  
  flows_2024 <- state_to_state_migration_2024()
  flows_2023 <- state_to_state_migration_2023()
  
  bind_rows(flows_2024, flows_2023)
}
```

```{r}
# Task 5
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(jsonlite)
library(stringr)

move_row_to_colnames <- function(X, row = 1) {
  X <- as_tibble(X)
  Xrow <- X[row, ]
  X <- X[-row, ]
  colnames(X) <- unlist(Xrow)
  readr::type_convert(X)
}

pull_state_from_metro <- function(metro_name) {
  str_extract(
    metro_name,
    ".*, (\\S{2})[-[:alpha:]]* Metro Area",
    group = 1
  )
}

metro_to_metro_migration <- function() {
  
  if (!dir.exists(file.path("data", "mp03"))) {
    dir.create(file.path("data", "mp03"), showWarnings = FALSE, recursive = TRUE)
  }
  
  json_file <- file.path(
    "data",
    "mp03",
    "metro_to_metro_migration_2020.json"
  )
  
  api_url <- paste0(
    "https://api.census.gov/data/2020/acs/flows",
    "?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME",
    "&for=metropolitan%20statistical%20area/micropolitan%20statistical%20area:*"
  )
  
  if (!file.exists(json_file)) {
    download.file(
      url = api_url,
      destfile = json_file,
      mode = "wb",
      quiet = TRUE
    )
  }
  
  metro_data <- read_json(
    json_file,
    simplifyVector = TRUE
  ) |>
    move_row_to_colnames() |>
    select(
      metro_current = FULL1_NAME,
      metro_1y = FULL2_NAME,
      moved_in = MOVEDIN,
      moved_out = MOVEDOUT
    ) |>
    mutate(
      metro_current_state = pull_state_from_metro(metro_current),
      metro_1y_state = pull_state_from_metro(metro_1y)
    )
  
  metro_data
}

```


```{r}
metro_flows <- metro_to_metro_migration()

dim(metro_flows)
head(metro_flows)

```


## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 




## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 



------------------------------------------------------------------------

This work ©2025 by KevMarcel0 was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
