---
title: "Mini-Project 04 - Going for the Gold"
author: "kevmarcel0"
editor:
    mode: source
format:
    html:
        code-fold: true
        toc: true
---

## Introduction

With the 2026 Winter Olympics complete and the 2028 Summer Olympics coming to Los
Angeles, serious money is being wagered on Team USA's performance. This report
builds a three-factor model to predict how many gold medals Team USA will win
in 2028, drawing on historical Olympic data scraped from
[Olympedia](https://www.olympedia.org/).

---

## Data Acquisition and Preparation


```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(rvest)
library(fs)
library(broom)
library(gt)
library(scales)
library(ggrepel)

# Rate-limited download (as provided by instructor)
slow_download <- purrr::slowly(download.file, purrr::rate_delay(pause = 5))

read_olympedia <- function(url, ...) {
  if (stringr::str_starts(url, "/")) url <- stringr::str_remove(url, "/")
  cache_dir <- fs::path("data", "mp04", "olympedia_cache")
  if (!fs::dir_exists(cache_dir)) fs::dir_create(cache_dir)

  cache_path <- fs::path(
    cache_dir,
    url |> as.character() |> stringr::str_replace_all("/", "_")
  )

  if (!fs::file_exists(cache_path)) {
    src_url <- paste0("https://www.olympedia.org/", url)
    slow_download(src_url, cache_path, mode = "wb")
  }

  rvest::read_html(cache_path)
}
```

### Step 1 — Parse All Olympic Editions

We scrape the editions index at `/editions`, extracting the edition ID from the
year-link href, the olympiad number from the Roman numeral column, the host
nation NOC from the flag image URL, and the season (Summer / Winter) from the
games-name text.

```{r}
#| warning: false
#| message: false

parse_editions <- function() {
  page   <- read_olympedia("editions")
  tables <- page |> html_elements("table")

  parse_season_table <- function(tbl, season_label) {
    rows <- tbl |> html_elements("tr")
    bind_rows(lapply(rows, function(row) {
      cells <- row |> html_elements("td")
      if (length(cells) < 4) return(NULL)

      # Cell 1: Roman numeral + href containing edition_id
      cell1_link <- cells[[1]] |> html_elements("a")
      if (length(cell1_link) == 0) return(NULL)

      href           <- cell1_link[[1]] |> html_attr("href")
      edition_id     <- str_extract(href, "\\d+$") |> as.integer()
      olympiad_roman <- cell1_link[[1]] |> html_text(trim = TRUE)

      # Cell 2: Year
      year <- cells[[2]] |> html_text(trim = TRUE) |> as.integer()

      # Cell 4: Host flag (.svg URL)
      flag_img <- cells[[4]] |> html_elements("img")
      host_noc <- if (length(flag_img) > 0) {
        str_extract(flag_img[[1]] |> html_attr("src"), "[A-Z]{3}(?=\\.svg)")
      } else {
        NA_character_
      }

      data.frame(
        edition_id     = edition_id,
        year           = year,
        olympiad_roman = olympiad_roman,
        host_noc       = host_noc,
        season         = season_label,
        stringsAsFactors = FALSE
      )
    }))
  }

  bind_rows(
    parse_season_table(tables[[1]], "Summer"),
    parse_season_table(tables[[2]], "Winter")
  ) |>
    as_tibble() |>
    filter(!is.na(edition_id), !is.na(year)) |>
    mutate(olympiad_num = suppressWarnings(as.integer(as.roman(olympiad_roman))))
}

editions <- parse_editions()

editions |>
  filter(season == "Summer") |>
  select(edition_id, year, olympiad_num, host_noc, season) |>
  tail(10) |>
  gt() |>
  tab_header(title = "Most Recent Summer Olympic Editions")
```

### Step 2 — Sports (Medal Disciplines) per Edition

For each edition we visit `/editions/{id}` and collect every link matching
the pattern `/editions/{id}/sports/{CODE}`.

```{r}
#| warning: false
#| message: false

parse_edition_sports <- function(edition_id) {
  empty_tbl <- tibble(
    sport_code = character(),
    sport_name = character(),
    sport_url  = character()
  )

  page <- tryCatch(
    read_olympedia(paste0("editions/", edition_id)),
    error = function(e) NULL
  )
  if (is.null(page)) return(empty_tbl)

  selector    <- sprintf("a[href*='/editions/%d/sports/']", edition_id)
  sport_nodes <- page |> html_elements(selector)

  if (length(sport_nodes) == 0) return(empty_tbl)

  tibble(
    sport_url  = sport_nodes |> html_attr("href"),
    sport_name = sport_nodes |> html_text(trim = TRUE)
  ) |>
    distinct(sport_url, .keep_all = TRUE) |>
    mutate(sport_code = str_extract(sport_url, "[A-Z0-9]+$")) |>
    filter(!is.na(sport_code), sport_code != "") |>
    select(sport_code, sport_name, sport_url)
}

all_sports <- editions |>
  mutate(sports = map(edition_id, parse_edition_sports)) |>
  unnest(sports)

cat(sprintf("Total edition × sport combinations: %s\n",
            comma(nrow(all_sports))))
```

### Step 3 — Medal Results per Sport × Edition

For each sport page we find the first table containing Olympedia flag images
(flag URLs like `...olympedia-flags.../USA.png`). We then parse each row as an
event record with Gold / Silver / Bronze columns, extracting the three-letter
NOC from each flag image src.

```{r}
#| warning: false
#| message: false

# Pull 3-letter NOC out of the first flag image in a table cell
extract_noc <- function(cell) {
  imgs <- cell |> html_elements("img[src*='olympedia-flags']")
  if (length(imgs) == 0) return(NA_character_)
  str_extract(imgs[[1]] |> html_attr("src"), "[A-Z]{3}(?=\\.(png|svg))")
}

parse_sport_medals <- function(edition_id, sport_code) {
  empty_tbl <- tibble(
    noc    = character(),
    gold   = integer(),
    silver = integer(),
    bronze = integer()
  )

  url  <- paste0("editions/", edition_id, "/sports/", sport_code)
  page <- tryCatch(read_olympedia(url), error = function(e) NULL)
  if (is.null(page) || !inherits(page, "xml_document")) return(empty_tbl)

  # Find the first table that contains flag images (the results table)
  tables      <- page |> html_elements("table")
  results_tbl <- NULL
  for (tbl in tables) {
    if (length(tbl |> html_elements("img[src*='olympedia-flags']")) > 0) {
      results_tbl <- tbl
      break
    }
  }
  if (is.null(results_tbl)) return(empty_tbl)

  # Use tbody rows to skip header automatically
  rows <- results_tbl |> html_elements("tbody tr")
  if (length(rows) == 0) {
    rows <- results_tbl |> html_elements("tr")
    if (length(rows) > 1) rows <- rows[-1]  # drop header
  }
  if (length(rows) == 0) return(empty_tbl)

  raw <- map_dfr(rows, function(row) {
    cells <- row |> html_elements("td")
    if (length(cells) < 2) return(NULL)

    tibble(
      event      = cells[[1]] |> html_text(trim = TRUE),
      gold_noc   = if (length(cells) >= 3) extract_noc(cells[[3]]) else NA_character_,
      silver_noc = if (length(cells) >= 5) extract_noc(cells[[5]]) else NA_character_,
      bronze_noc = if (length(cells) >= 7) extract_noc(cells[[7]]) else NA_character_
    )
  })

  if (nrow(raw) == 0) return(empty_tbl)

  # Pivot → long → aggregate by NOC and medal type
  raw |>
    pivot_longer(
      cols      = ends_with("_noc"),
      names_to  = "medal_type",
      values_to = "noc"
    ) |>
    filter(!is.na(noc), noc != "") |>
    mutate(medal_type = str_remove(medal_type, "_noc")) |>
    count(noc, medal_type, name = "medals") |>
    pivot_wider(
      names_from  = medal_type,
      values_from = medals,
      values_fill = 0L
    )
}

# Scrape all sport × edition combinations
all_medals_raw <- all_sports |>
  select(edition_id, sport_code) |>
  distinct() |>
  mutate(data = map2(edition_id, sport_code, parse_sport_medals)) |>
  unnest(data)
```

### Step 4 — Assemble the Master Table

```{r}
olympic_medals <- all_medals_raw |>
  left_join(
    editions |> select(edition_id, year, season, host_noc, olympiad_num),
    by = "edition_id"
  ) |>
  left_join(
    all_sports |> select(edition_id, sport_code, sport_name) |> distinct(),
    by = c("edition_id", "sport_code")
  ) |>
  mutate(
    gold   = replace_na(gold,   0L),
    silver = replace_na(silver, 0L),
    bronze = replace_na(bronze, 0L),
    total  = gold + silver + bronze,
    is_host = (noc == host_noc)
  )

# Verify with a random sample
set.seed(42)
olympic_medals |>
  filter(total > 0, season == "Summer") |>
  slice_sample(n = 5) |>
  select(year, sport_name, noc, gold, silver, bronze, total) |>
  gt() |>
  tab_header(title = "Olympic Medal Results by Discipline (5 random rows)") |>
  cols_label(
    year = "Year", sport_name = "Discipline", noc = "NOC",
    gold = "Gold", silver = "Silver", bronze = "Bronze", total = "Total"
  )
```

---

## Task 2: Standardize and Clean Countries

Olympedia uses **National Olympic Committee (NOC) codes**, which frequently
differ from the ISO 3166-1 alpha-3 codes

### Building the NOC → ISO-3 Lookup

```{r}
# Comprehensive NOC → ISO-3 lookup
# Keys   = Olympedia NOC codes
# Values = ISO 3166-1 alpha-3 codes expected by necountries::countries()
noc_to_iso <- c(
  # Europe
  "GRE" = "GRC",  # Greece
  "GER" = "DEU",  # Germany
  "DEN" = "DNK",  # Denmark
  "NED" = "NLD",  # Netherlands
  "POR" = "PRT",  # Portugal
  "SUI" = "CHE",  # Switzerland
  "CRO" = "HRV",  # Croatia
  "SLO" = "SVN",  # Slovenia
  "BUL" = "BGR",  # Bulgaria
  "ROM" = "ROU",  # Romania
  "LAT" = "LVA",  # Latvia
  "MON" = "MCO",  # Monaco

  # Asia / Middle East
  "IRI" = "IRN",  # Iran
  "MAS" = "MYS",  # Malaysia
  "SIN" = "SGP",  # Singapore
  "TPE" = "TWN",  # Chinese Taipei
  "VIE" = "VNM",  # Vietnam
  "MGL" = "MNG",  # Mongolia
  "NEP" = "NPL",  # Nepal
  "SRI" = "LKA",  # Sri Lanka
  "BAN" = "BGD",  # Bangladesh

  # Africa
  "NGR" = "NGA",  # Nigeria
  "ZIM" = "ZWE",  # Zimbabwe
  "RSA" = "ZAF",  # South Africa
  "BOT" = "BWA",  # Botswana
  "TAN" = "TZA",  # Tanzania
  "ZAM" = "ZMB",  # Zambia
  "GUI" = "GIN",  # Guinea
  "MAW" = "MWI",  # Malawi
  "SUD" = "SDN",  # Sudan
  "SEY" = "SYC",  # Seychelles
  "TOG" = "TGO",  # Togo
  "ALG" = "DZA",  # Algeria
  "LBA" = "LBY",  # Libya
  "MRI" = "MUS",  # Mauritius
  "NIG" = "NER",  # Niger

  # Americas / Caribbean
  "HAI" = "HTI",  # Haiti
  "BAH" = "BHS",  # Bahamas
  "GUA" = "GTM",  # Guatemala
  "PHI" = "PHL",  # Philippines
  "HON" = "HND",  # Honduras
  "PAR" = "PRY",  # Paraguay
  "URU" = "URY",  # Uruguay
  "CHI" = "CHL",  # Chile
  "CRC" = "CRI",  # Costa Rica
  "ESA" = "SLV",  # El Salvador
  "NCA" = "NIC",  # Nicaragua
  "ANT" = "ATG",  # Antigua & Barbuda
  "PUR" = "PRI",  # Puerto Rico
  "VIN" = "VCT",  # St. Vincent
  "SKN" = "KNA",  # St. Kitts & Nevis

  # Pacific
  "TGA" = "TON",  # Tonga
  "SAM" = "WSM",  # Samoa
  "FIJ" = "FJI",  # Fiji

  # Historical / defunct — mapped to principal successor state
  "URS" = "RUS",  # Soviet Union        → Russia
  "EUN" = "RUS",  # Unified Team (1992) → Russia
  "TCH" = "CZE",  # Czechoslovakia      → Czech Republic
  "GDR" = "DEU",  # East Germany        → Germany
  "FRG" = "DEU",  # West Germany        → Germany
  "YUG" = "SRB",  # Yugoslavia          → Serbia
  "SCG" = "SRB",  # Serbia & Montenegro → Serbia
  "ANZ" = "AUS",  # Australasia         → Australia
  "BOH" = "CZE",  # Bohemia             → Czech Republic
  "WIF" = "TTO"   # West Indies Fed.    → Trinidad & Tobago
)

# Apply mapping; codes not in the lookup keep their original value
# replace_values() from recent dplyr serves the same purpose as recode() below
olympic_medals_std <- olympic_medals |>
  mutate(iso3 = dplyr::recode(noc, !!!noc_to_iso, .default = noc))

cat(sprintf(
  "Distinct NOC codes in raw data:        %d\n",
  n_distinct(olympic_medals$noc)
))
cat(sprintf(
  "Distinct ISO-3 codes after mapping:    %d\n",
  n_distinct(olympic_medals_std$iso3)
))
```

### Finding Remaining Mismatches with `anti_join()`

```{r}
library(necountries)
library(sf)

world <- countries()   # sf data frame with iso3, country, polygon geometry

# Pseudo-codes that legitimately have no country geometry
pseudo_codes <- c(
  "IOA",  # Individual Olympic Athletes
  "ROT",  # Refugee Olympic Team
  "EOR",  # Équipe Olympique des Réfugiés
  "ANO",  # Athletes of NOC (neutral)
  "IOP",  # Independent Olympic Participants
  "NFL",  # Newfoundland (pre-1949)
  "SAA"   # Saar (1952)
)

unmatched <- olympic_medals_std |>
  distinct(noc, iso3) |>
  anti_join(world, by = "iso3") |>
  filter(!iso3 %in% pseudo_codes) |>
  arrange(iso3)

cat(sprintf("%d NOC codes still unmatched after initial mapping:\n\n",
            nrow(unmatched)))
print(unmatched, n = Inf)
```

```{r}
# Spot-fix any residual codes surfaced by anti_join above
# (populate as needed from the unmatched table)
spot_fixes <- c(
  "KOS" = "XKX",  # Kosovo (ISO uses XKX as unofficial code)
  "SGP" = "SGP"   # placeholder — add real fixes here after inspecting output
)

if (length(spot_fixes) > 0) {
  olympic_medals_std <- olympic_medals_std |>
    mutate(iso3 = dplyr::recode(iso3, !!!spot_fixes, .default = iso3))
}

# Final check — only pseudo-codes should remain unmatched
final_unmatched <- olympic_medals_std |>
  distinct(noc, iso3) |>
  anti_join(world, by = "iso3") |>
  filter(!iso3 %in% pseudo_codes)

cat(sprintf("%d unresolvable codes remain (pseudo/historical teams).\n",
            nrow(final_unmatched)))
```

### Join Olympic Data to `countries()` sf Object

```{r}
# All-time Summer medal totals per standardised ISO-3 code
medals_by_country <- olympic_medals_std |>
  filter(season == "Summer") |>
  group_by(iso3) |>
  summarise(
    total_gold   = sum(gold),
    total_silver = sum(silver),
    total_bronze = sum(bronze),
    games_attended = n_distinct(year),
    .groups = "drop"
  )

# Left-join so every country in the map is retained
world_medals <- world |>
  left_join(medals_by_country, by = "iso3") |>
  mutate(
    total_gold     = replace_na(total_gold,   0),
    total_silver   = replace_na(total_silver, 0),
    total_bronze   = replace_na(total_bronze, 0),
    gold_per_game  = if_else(
      games_attended > 0,
      total_gold / games_attended,
      NA_real_
    )
  )
```

### Geospatial Visualization 1 — All-Time Gold Medals

```{r}
#| fig-width: 12
#| fig-height: 6

ggplot(world_medals) +
  geom_sf(aes(fill = total_gold), color = "white", linewidth = 0.12) +
  scale_fill_gradient(
    low      = "#FFF9C4",
    high     = "#8B6914",
    trans    = "log1p",
    breaks   = c(0, 1, 5, 25, 100, 500, 1000),
    labels   = scales::comma,
    na.value = "grey85",
    name     = "Gold Medals\n(log scale)"
  ) +
  labs(
    title    = "All-Time Summer Olympic Gold Medals by Country",
    subtitle = "Historical NOC codes standardised to ISO-3; defunct nations mapped to successors",
    caption  = "Source: Olympedia.org"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40", size = 11),
    legend.position  = "bottom",
    legend.key.width = unit(2.5, "cm")
  )
```

### Geospatial Visualization 2 — Consistency: Golds per Games Attended

```{r}
#| fig-width: 12
#| fig-height: 6

ggplot(world_medals) +
  geom_sf(aes(fill = gold_per_game), color = "white", linewidth = 0.12) +
  scale_fill_viridis_c(
    option   = "plasma",
    trans    = "sqrt",
    na.value = "grey85",
    name     = "Golds per\nGames Attended",
    labels   = scales::number_format(accuracy = 0.1)
  ) +
  labs(
    title    = "Consistent Olympic Powerhouses — Gold Medals per Games Attended",
    subtitle = "Controlling for participation history surfaces Cuba, Hungary, Finland alongside the usual giants",
    caption  = "Source: Olympedia.org"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 14),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40", size = 11),
    legend.position  = "bottom",
    legend.key.width = unit(2.5, "cm")
  )
```

---

## Exploratory Analysis

### USA's Gold Medal History at the Summer Olympics

```{r}
#| fig-width: 10
#| fig-height: 5

# Per-games totals for USA and all countries
games_totals <- olympic_medals |>
  filter(season == "Summer") |>
  group_by(year, edition_id, host_noc) |>
  summarise(
    total_events = sum(gold),   # total golds ≈ total events
    .groups = "drop"
  )

usa_by_year <- olympic_medals |>
  filter(season == "Summer", noc == "USA") |>
  group_by(year) |>
  summarise(usa_gold = sum(gold), .groups = "drop")

usa_summary <- games_totals |>
  left_join(usa_by_year, by = "year") |>
  replace_na(list(usa_gold = 0)) |>
  mutate(
    usa_share  = usa_gold / total_events,
    is_usa_host = (host_noc == "USA")
  )

ggplot(usa_summary, aes(x = year, y = usa_gold)) +
  geom_line(color = "grey70", linewidth = 0.8) +
  geom_point(aes(color = is_usa_host), size = 3) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "gold2"),
    labels = c("Away", "Home (USA hosts)")
  ) +
  labs(
    title   = "Team USA — Summer Olympic Gold Medals",
    subtitle = "Gold dots mark years when the US hosted",
    x = "Year", y = "Gold Medals", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

```{r}
#| fig-width: 10
#| fig-height: 4

ggplot(usa_summary, aes(x = year, y = usa_share)) +
  geom_line(color = "grey70", linewidth = 0.8) +
  geom_point(aes(color = is_usa_host), size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "gold2"),
    labels = c("Away", "Home")
  ) +
  labs(
    title    = "Team USA — Share of All Summer Olympic Gold Medals",
    subtitle = "Using share controls for the growing number of events over time",
    x = "Year", y = "Gold Medal Share", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```


USA dominated early games (1904 St. Louis — home advantage + limited participation) and 1984 Los Angeles (Soviet-bloc boycott) — both outliers. With clear spikes are visible in 1904, 1932, 1984, and 1996 — all years the USA
hosted.

---

## Task 3: Exploratory Data Analysis

### US Baseline Performance

##1

```{r}
#| echo: false
total_usa_gold <- olympic_medals |>
  filter(season == "Summer", noc == "USA") |>
  summarise(n = sum(gold)) |>
  pull(n)

avg_usa_gold_modern <- olympic_medals |>
  filter(season == "Summer", noc == "USA", year >= 1992, year <= 2024) |>
  group_by(year) |>
  summarise(g = sum(gold), .groups = "drop") |>
  summarise(avg = mean(g)) |>
  pull(avg)
```

Team USA has won `r total_usa_gold` gold medals across all Summer Olympics and averages `r round(avg_usa_gold_modern, 1)` golds per games in the modern era (1992–2024).

##2

```{r}
#| echo: false
olympic_medals |>
  filter(season == "Summer", noc == "USA", year >= 1960) |>
  mutate(decade = paste0(floor(year / 10) * 10, "s")) |>
  group_by(decade) |>
  summarise(
    games    = n_distinct(year),
    gold     = sum(gold),
    silver   = sum(silver),
    bronze   = sum(bronze),
    total    = sum(total),
    avg_gold = round(gold / games, 1),
    .groups  = "drop"
  ) |>
  gt() |>
  tab_header(title = "Team USA — Summer Olympic Medals by Decade (1960s onward)") |>
  cols_label(
    decade = "Decade", games = "Games Played", gold = "Gold",
    silver = "Silver", bronze = "Bronze", total = "Total",
    avg_gold = "Avg Gold / Games"
  )
```

##3

```{r}
#| echo: false
#| warning: false
top5_countries <- olympic_medals |>
  filter(season == "Summer", year >= 1960) |>
  group_by(noc) |>
  summarise(total_gold = sum(gold), .groups = "drop") |>
  arrange(desc(total_gold)) |>
  slice_head(n = 5) |>
  pull(noc)

olympic_medals |>
  filter(season == "Summer", year >= 1960, noc %in% top5_countries) |>
  group_by(year, noc) |>
  summarise(gold = sum(gold), .groups = "drop") |>
  ggplot(aes(x = year, y = gold, color = noc)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  labs(
    title    = "Gold Medals — Top 5 Countries at Summer Olympics (1960–2024)",
    subtitle = "URS/EUN medals consolidated under RUS for continuity",
    x = "Year", y = "Gold Medals", color = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The US lead is clear but has narrowed since the Cold War era.

---

### Host Country Effect

##4

```{r}
#| echo: false
# For each Summer Olympics, compare host's medals to prior same-season Olympics
country_medals_yr <- olympic_medals |>
  filter(season == "Summer") |>
  group_by(year, noc) |>
  summarise(
    gold = sum(gold), silver = sum(silver),
    bronze = sum(bronze), total = sum(total),
    .groups = "drop"
  )

summer_year_order <- editions |>
  filter(season == "Summer") |>
  arrange(year) |>
  mutate(prev_year = lag(year)) |>
  select(year, host_noc, prev_year)

host_comparison <- summer_year_order |>
  filter(!is.na(host_noc), !is.na(prev_year), year >= 1960) |>
  left_join(
    country_medals_yr |>
      rename(curr_gold = gold, curr_silver = silver,
             curr_bronze = bronze, curr_total = total),
    by = c("year", "host_noc" = "noc")
  ) |>
  left_join(
    country_medals_yr |>
      rename(prev_gold = gold, prev_silver = silver,
             prev_bronze = bronze, prev_total = total),
    by = c("prev_year" = "year", "host_noc" = "noc")
  ) |>
  filter(!is.na(curr_gold), !is.na(prev_gold), prev_gold > 0) |>
  mutate(
    pct_gold   = (curr_gold   - prev_gold)   / prev_gold   * 100,
    pct_silver = (curr_silver - prev_silver) / prev_silver * 100,
    pct_bronze = (curr_bronze - prev_bronze) / prev_bronze * 100,
    pct_total  = (curr_total  - prev_total)  / prev_total  * 100
  )

med_host_pct <- round(median(host_comparison$pct_gold, na.rm = TRUE), 1)
n_improved   <- sum(host_comparison$pct_gold > 0, na.rm = TRUE)
n_total_host <- nrow(host_comparison)
```

Of `r n_total_host` hosting occasions since 1960, `r n_improved` saw the host improve their gold count vs the prior Summer Olympics. The median change is `r med_host_pct`%.

##5

```{r}
#| echo: false
host_comparison |>
  select(year, host_noc, prev_year, prev_gold, curr_gold, pct_gold) |>
  arrange(year) |>
  mutate(pct_gold = round(pct_gold, 1)) |>
  gt() |>
  tab_header(title = "Host Country Gold Medals: Hosting Year vs Prior Summer Olympics") |>
  cols_label(
    year = "Host Year", host_noc = "Host",
    prev_year = "Prior Year", prev_gold = "Prior Gold",
    curr_gold = "Host-Year Gold", pct_gold = "% Change"
  ) |>
  tab_style(
    style     = cell_fill(color = "#d4edda"),
    locations = cells_body(rows = pct_gold > 0)
  ) |>
  tab_style(
    style     = cell_fill(color = "#f8d7da"),
    locations = cells_body(rows = pct_gold <= 0)
  )
```

Green rows improved; red rows declined. The 1984 LA spike reflects the Soviet boycott.

##6

```{r}
#| echo: false
host_comparison |>
  ggplot(aes(x = prev_gold, y = curr_gold, label = paste0(host_noc, "\n", year))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = pct_gold > 0), size = 3) +
  ggrepel::geom_text_repel(size = 2.8, max.overlaps = 8) +
  scale_color_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "tomato"),
    labels = c("TRUE" = "Improved", "FALSE" = "Declined")
  ) +
  labs(
    title    = "Host Country: Prior Summer Games Gold vs Host-Year Gold",
    subtitle = "Points above the dashed line improved; below declined",
    x = "Gold Medals (Prior Summer Olympics)",
    y = "Gold Medals (Host Year)",
    color = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

### New Sport Effect

##7

```{r}
#| echo: false
# New sports = sports not in the immediately preceding Summer Olympics
sports_per_year <- olympic_medals |>
  filter(season == "Summer") |>
  distinct(year, sport_code)

year_prev_map <- sports_per_year |>
  distinct(year) |>
  arrange(year) |>
  mutate(prev_year = lag(year))

new_sports_per_year <- sports_per_year |>
  inner_join(year_prev_map, by = "year") |>
  filter(!is.na(prev_year)) |>
  anti_join(
    sports_per_year |> rename(prev_year = year),
    by = c("prev_year", "sport_code")
  ) |>
  select(year, sport_code)

# For each new sport, did the host win gold / silver / bronze?
host_new_results <- olympic_medals |>
  filter(season == "Summer") |>
  inner_join(new_sports_per_year, by = c("year", "sport_code")) |>
  filter(!is.na(host_noc), year >= 1960) |>
  group_by(year, sport_code, sport_name, host_noc) |>
  summarise(
    host_gold   = sum(gold[noc   == host_noc[1]], na.rm = TRUE),
    host_silver = sum(silver[noc == host_noc[1]], na.rm = TRUE),
    host_bronze = sum(bronze[noc == host_noc[1]], na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    won_gold   = host_gold   > 0,
    won_silver = host_silver > 0,
    won_bronze = host_bronze > 0,
    won_any    = (host_gold + host_silver + host_bronze) > 0
  )

n_debut_sports  <- nrow(host_new_results)
host_gold_rate  <- round(mean(host_new_results$won_gold)  * 100, 1)
host_medal_rate <- round(mean(host_new_results$won_any)   * 100, 1)
```

Since 1960 there have been `r n_debut_sports` debut sport appearances. The host won gold in `r host_gold_rate`% of them and at least one medal in `r host_medal_rate`%.

##8

```{r}
#| echo: false
host_new_results |>
  arrange(year) |>
  select(year, host_noc, sport_name, host_gold, host_silver, host_bronze, won_gold) |>
  gt() |>
  tab_header(title = "Host Nation Medals in Debut Sports (1960 – 2024)") |>
  cols_label(
    year = "Year", host_noc = "Host", sport_name = "New Sport",
    host_gold = "Gold", host_silver = "Silver", host_bronze = "Bronze",
    won_gold = "Won Gold?"
  ) |>
  tab_style(
    style     = cell_fill(color = "#FFF3CD"),
    locations = cells_body(rows = won_gold == TRUE)
  )
```

##9

```{r}
#| echo: false
# Win rate for host in non-debut sports
host_established <- olympic_medals |>
  filter(season == "Summer", year >= 1960) |>
  anti_join(new_sports_per_year, by = c("year", "sport_code")) |>
  filter(!is.na(host_noc)) |>
  group_by(year, sport_code, host_noc) |>
  summarise(
    host_gold = sum(gold[noc == host_noc[1]], na.rm = TRUE),
    won_gold  = host_gold > 0,
    .groups = "drop"
  )

win_rate_df <- bind_rows(
  host_new_results      |> mutate(type = "Debut Sport"),
  host_established      |> mutate(type = "Established Sport")
) |>
  group_by(type) |>
  summarise(
    win_rate = mean(won_gold, na.rm = TRUE),
    n        = n(),
    se       = sqrt(win_rate * (1 - win_rate) / n),
    .groups  = "drop"
  )

ggplot(win_rate_df, aes(x = type, y = win_rate, fill = type)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_errorbar(
    aes(ymin = win_rate - 1.96 * se, ymax = win_rate + 1.96 * se),
    width = 0.2
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 0.5)) +
  scale_fill_manual(values = c("Debut Sport" = "gold2", "Established Sport" = "steelblue")) +
  labs(
    title    = "Host Nation Gold-Win Rate: Debut vs Established Sports",
    subtitle = "Error bars show 95% CI. Host nations win gold at higher rates in sports they introduce.",
    x = NULL, y = "Probability of Winning Gold"
  ) +
  theme_minimal()
```

---

### Additional Findings

##10

```{r}
#| echo: false
usa_home <- usa_summary |>
  filter(is_usa_host) |>
  summarise(avg = mean(usa_gold)) |>
  pull(avg)

usa_away <- usa_summary |>
  filter(!is_usa_host, year >= 1960) |>
  summarise(avg = mean(usa_gold)) |>
  pull(avg)
```

When the US hosts, it averages **`r round(usa_home, 1)`** gold medals per
games. In away games since 1960, that average is **`r round(usa_away, 1)`** —
a home advantage of roughly **`r round(usa_home - usa_away, 1)`** extra golds.

##11

```{r}
#| echo: false
olympic_medals |>
  filter(season == "Summer") |>
  group_by(noc) |>
  summarise(
    gold   = sum(gold),
    silver = sum(silver),
    bronze = sum(bronze),
    total  = sum(total),
    .groups = "drop"
  ) |>
  arrange(desc(gold)) |>
  slice_head(n = 10) |>
  gt() |>
  tab_header(title = "All-Time Top 10 Countries — Summer Olympic Gold Medals") |>
  cols_label(
    noc = "Country (NOC)", gold = "Gold", silver = "Silver",
    bronze = "Bronze", total = "Total"
  )
```

##12

```{r}
#| echo: false
games_totals |>
  ggplot(aes(x = year, y = total_events)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "red", linetype = "dashed") +
  labs(
    title    = "Total Medal Events at Summer Olympics Over Time",
    subtitle = "Steady growth in number of events inflates raw medal counts",
    x = "Year", y = "Total Gold Medals Awarded"
  ) +
  theme_minimal()
```

The program has grown from 43 events in 1896 to over 300 in 2024, making raw counts misleading without normalizing by share.

---

## Task 4: Statistical Inference

```{r}
#| echo: false
#| warning: false
#| message: false
library(infer)
```

##Q1

95% CIs for Team USA's per-games medal count (Summer Olympics, 1960–2024) from one-sample t-tests.

```{r}
#| echo: false
usa_medals_yr <- olympic_medals |>
  filter(season == "Summer", noc == "USA", year >= 1960) |>
  group_by(year) |>
  summarise(
    gold   = sum(gold),
    silver = sum(silver),
    bronze = sum(bronze),
    total  = sum(total),
    .groups = "drop"
  )

usa_ci <- bind_rows(
  usa_medals_yr |> t_test(response = gold)   |> mutate(medal = "Gold"),
  usa_medals_yr |> t_test(response = silver) |> mutate(medal = "Silver"),
  usa_medals_yr |> t_test(response = bronze) |> mutate(medal = "Bronze"),
  usa_medals_yr |> t_test(response = total)  |> mutate(medal = "Total")
)

usa_ci |>
  select(medal, estimate, lower_ci, upper_ci) |>
  mutate(across(c(estimate, lower_ci, upper_ci), \(x) round(x, 1))) |>
  gt() |>
  tab_header(title = "Team USA — Average Medals per Summer Olympics (1960–2024)",
             subtitle = "95% confidence intervals from one-sample t-test") |>
  cols_label(medal = "Medal Type", estimate = "Estimate",
             lower_ci = "95% CI Lower", upper_ci = "95% CI Upper")
```

##Q2

Median host gold improvement vs the prior same-season Olympics is `r med_host_pct`% — both the % change CI and the proportion of improved occasions are tested.

```{r}
#| echo: false
host_ci <- bind_rows(
  host_comparison |> t_test(response = pct_gold)   |> mutate(medal = "Gold"),
  host_comparison |> t_test(response = pct_silver) |> mutate(medal = "Silver"),
  host_comparison |> t_test(response = pct_bronze) |> mutate(medal = "Bronze"),
  host_comparison |> t_test(response = pct_total)  |> mutate(medal = "Total")
)

host_ci |>
  select(medal, estimate, lower_ci, upper_ci) |>
  mutate(across(c(estimate, lower_ci, upper_ci), \(x) round(x, 1))) |>
  gt() |>
  tab_header(
    title    = "Host Country % Change in Medals vs Prior Same-Season Olympics",
    subtitle = "95% CI from one-sample t-test. Positive = more medals when hosting."
  ) |>
  cols_label(medal = "Medal Type", estimate = "% Change Estimate",
             lower_ci = "95% CI Lower", upper_ci = "95% CI Upper")
```

```{r}
#| echo: false
n_improved <- sum(host_comparison$pct_gold > 0, na.rm = TRUE)
n_total    <- sum(!is.na(host_comparison$pct_gold))
bt <- binom.test(n_improved, n_total, alternative = "greater")

tibble(
  improved    = n_improved,
  total       = n_total,
  proportion  = round(n_improved / n_total, 3),
  p_value     = round(bt$p.value, 4),
  ci_lower    = round(bt$conf.int[1], 3),
  ci_upper    = round(bt$conf.int[2], 3)
) |>
  gt() |>
  tab_header(
    title    = "Proportion of Hosting Occasions Where Host Improved Gold Count",
    subtitle = "One-sided binomial test (H1: p > 0.5)"
  ) |>
  cols_label(
    improved = "Improved", total = "Total", proportion = "Proportion",
    p_value = "p-value", ci_lower = "95% CI Lower", ci_upper = "95% CI Upper"
  )
```

##Q3

Proportion test CI for the probability the host wins gold in a debut sport (1960–2024).

```{r}
#| echo: false
new_sport_ci <- bind_rows(
  host_new_results |>
    prop_test(won_gold   ~ NULL) |> mutate(medal = "Gold"),
  host_new_results |>
    prop_test(won_silver ~ NULL) |> mutate(medal = "Silver"),
  host_new_results |>
    prop_test(won_bronze ~ NULL) |> mutate(medal = "Bronze"),
  host_new_results |>
    mutate(won_any = as.logical(won_any)) |>
    prop_test(won_any ~ NULL) |> mutate(medal = "Any Medal")
)

new_sport_ci |>
  select(medal, lower_ci, upper_ci) |>
  mutate(
    estimate = (lower_ci + upper_ci) / 2,
    across(c(estimate, lower_ci, upper_ci), \(x) round(x, 3))
  ) |>
  select(medal, estimate, lower_ci, upper_ci) |>
  gt() |>
  tab_header(
    title    = "Probability Host Wins Medal in a Debut Sport",
    subtitle = "95% CI from one-sample proportion test (1960 – 2024)"
  ) |>
  cols_label(medal = "Medal Type", estimate = "Probability Estimate",
             lower_ci = "95% CI Lower", upper_ci = "95% CI Upper")
```

---

## Task 5: Combining Confidence Intervals

We forecast Team USA's 2028 gold medals using:

$$\text{USA}_{2028} = \underbrace{B}_{\text{baseline}} \times \left(1 + \frac{H}{100}\right) + P \times 22$$

where $B$ is USA's average gold count per games (Task 4 Q1), $H$ is the host nation's percent increase over the prior same-season Olympics (Task 4 Q2), $P$ is the probability of winning gold in a debut sport (Task 4 Q3), and **22** is the additional medal events in 2028 vs 2024. We assume the three intervals are independent and combine them via Monte Carlo simulation.

```{r}
#| echo: false
set.seed(2028)
n_sim <- 1e7

# ── 1. Baseline: avg USA gold medals ──────────────────────────
b_row       <- usa_ci |> filter(medal == "Gold")
b_ci_lower  <- b_row$lower_ci
b_ci_upper  <- b_row$upper_ci
b_ci_mean   <- (b_ci_lower + b_ci_upper) / 2
b_ci_sd     <- (b_ci_upper - b_ci_mean) / qnorm(1 - 0.05 / 2)

# ── 2. Host pct increase ──────────────────────────────────────
h_row       <- host_ci |> filter(medal == "Gold")
h_ci_lower  <- h_row$lower_ci
h_ci_upper  <- h_row$upper_ci
h_ci_mean   <- (h_ci_lower + h_ci_upper) / 2
h_ci_sd     <- (h_ci_upper - h_ci_mean) / qnorm(1 - 0.05 / 2)

# ── 3. Prob of winning gold in a debut sport ──────────────────
n_row       <- new_sport_ci |> filter(medal == "Gold")
n_ci_lower  <- n_row$lower_ci
n_ci_upper  <- n_row$upper_ci
n_ci_mean   <- (n_ci_lower + n_ci_upper) / 2
n_ci_sd     <- (n_ci_upper - n_ci_mean) / qnorm(1 - 0.05 / 2)

# ── Monte Carlo ───────────────────────────────────────────────
b_samples <- rnorm(n_sim, mean = b_ci_mean, sd = b_ci_sd)
h_samples <- rnorm(n_sim, mean = h_ci_mean, sd = h_ci_sd)
p_samples <- pmax(0, pmin(1, rnorm(n_sim, mean = n_ci_mean, sd = n_ci_sd)))

gold_2028_samples  <- b_samples * (1 + h_samples / 100) + p_samples * 22
gold_2028_interval <- quantile(gold_2028_samples, c(0.025, 0.975))
gold_2028_mean     <- mean(gold_2028_samples)
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 4
ggplot(tibble(gold = gold_2028_samples[gold_2028_samples < quantile(gold_2028_samples, 0.999)]),
       aes(x = gold)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 70, fill = "gold2", alpha = 0.85, color = "white") +
  geom_density(color = "navy", linewidth = 1.1) +
  geom_vline(xintercept = gold_2028_mean, color = "red",
             linetype = "dashed", linewidth = 1.2) +
  annotate("text", x = gold_2028_mean + 2, y = Inf, vjust = 2,
           label = sprintf("Mean: %.0f", gold_2028_mean),
           color = "red", fontface = "bold", size = 4.5) +
  labs(
    title    = "Monte Carlo Forecast — Team USA Gold Medals at LA 2028",
    subtitle = sprintf("10,000,000 simulations  |  95%% CI: [%.0f, %.0f]",
                       gold_2028_interval[1], gold_2028_interval[2]),
    x = "Predicted Gold Medals", y = "Density"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| echo: false
simulate_2028 <- function(medal_type) {
  b  <- usa_ci  |> filter(medal == medal_type)
  h  <- host_ci |> filter(medal == medal_type)
  ns <- new_sport_ci |> filter(medal == medal_type)

  b_mean <- (b$lower_ci  + b$upper_ci)  / 2
  h_mean <- (h$lower_ci  + h$upper_ci)  / 2
  p_mean <- (ns$lower_ci + ns$upper_ci) / 2

  b_sd <- (b$upper_ci  - b_mean)  / qnorm(1 - 0.05 / 2)
  h_sd <- (h$upper_ci  - h_mean)  / qnorm(1 - 0.05 / 2)
  p_sd <- (ns$upper_ci - p_mean)  / qnorm(1 - 0.05 / 2)

  b_s <- rnorm(n_sim, b_mean, b_sd)
  h_s <- rnorm(n_sim, h_mean, h_sd)
  p_s <- pmax(0, pmin(1, rnorm(n_sim, p_mean, p_sd)))

  sims <- pmax(0, b_s * (1 + h_s / 100) + p_s * 22)
  tibble(
    `Medal Type` = medal_type,
    `Point Estimate` = round(mean(sims), 1),
    `95% CI` = sprintf("[%.0f, %.0f]", quantile(sims, 0.025), quantile(sims, 0.975))
  )
}

set.seed(2028)
bind_rows(
  simulate_2028("Gold"),
  simulate_2028("Silver"),
  simulate_2028("Bronze")
) |>
  gt() |>
  tab_header(title = "Task 5 — Team USA 2028 Medal Forecast",
             subtitle = "Monte Carlo (10M simulations), 95% CI assumes independent factors") |>
  tab_style(
    style     = list(cell_fill(color = "#FFF3CD"), cell_text(weight = "bold")),
    locations = cells_body(rows = 1)
  )
```

---

## Task 6: Team USA 2028 — Fundraising Fact Sheet




```{r}
#| echo: false
tibble(
  Metric = c("Predicted Gold Medals", "95% Confidence Range", "Predicted Silver Medals", "Predicted Bronze Medals"),
  Value  = c(
    sprintf("%.0f", gold_2028_mean),
    sprintf("[%.0f – %.0f]", gold_2028_interval[1], gold_2028_interval[2]),
    sprintf("%.0f", {
      set.seed(2028)
      b <- usa_ci |> filter(medal == "Silver")
      h <- host_ci |> filter(medal == "Silver")
      ns <- new_sport_ci |> filter(medal == "Silver")
      b_m <- (b$lower_ci+b$upper_ci)/2; h_m <- (h$lower_ci+h$upper_ci)/2; p_m <- (ns$lower_ci+ns$upper_ci)/2
      mean(pmax(0, rnorm(1e6,b_m,(b$upper_ci-b_m)/qnorm(0.975))*(1+rnorm(1e6,h_m,(h$upper_ci-h_m)/qnorm(0.975))/100)+pmax(0,pmin(1,rnorm(1e6,p_m,(ns$upper_ci-p_m)/qnorm(0.975))))*22))
    }),
    sprintf("%.0f", {
      b <- usa_ci |> filter(medal == "Bronze")
      h <- host_ci |> filter(medal == "Bronze")
      ns <- new_sport_ci |> filter(medal == "Bronze")
      b_m <- (b$lower_ci+b$upper_ci)/2; h_m <- (h$lower_ci+h$upper_ci)/2; p_m <- (ns$lower_ci+ns$upper_ci)/2
      mean(pmax(0, rnorm(1e6,b_m,(b$upper_ci-b_m)/qnorm(0.975))*(1+rnorm(1e6,h_m,(h$upper_ci-h_m)/qnorm(0.975))/100)+pmax(0,pmin(1,rnorm(1e6,p_m,(ns$upper_ci-p_m)/qnorm(0.975))))*22))
    })
  )
) |>
  gt() |>
  tab_header(title = "Team USA — LA 2028 Medal Projection") |>
  tab_style(
    style     = list(cell_fill(color = "#003DA5"), cell_text(color = "white", weight = "bold")),
    locations = cells_body(rows = 1)
  )
```




Host nations consistently outperform their own prior results. Across all modern Olympics, the host country's gold medal count increases by an average of `r round(h_ci_mean, 1)`% compared to where they finished the *previous* Summer Games (95% CI: [`r round(h_ci_lower, 1)`%, `r round(h_ci_upper, 1)`%]). For the US at this scale, that translates to a meaningful uplift on top of the baseline.



### Why Sponsors Should Feel Confident

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.5
tibble(
  Factor = c("Baseline", "Home Field", "New Sports", "Total (Monte Carlo)"),
  Low    = c(b_ci_lower, b_ci_lower*(1+h_ci_lower/100), n_ci_lower*22, gold_2028_interval[1]),
  Mid    = c(b_ci_mean,  b_ci_mean*(1+h_ci_mean/100),   n_ci_mean*22,  gold_2028_mean),
  High   = c(b_ci_upper, b_ci_upper*(1+h_ci_upper/100), n_ci_upper*22, gold_2028_interval[2])
) |>
  mutate(Factor = factor(Factor, levels = rev(c("Baseline","Home Field","New Sports","Total (Monte Carlo)")))) |>
  ggplot(aes(y = Factor)) +
  geom_segment(aes(x = Low, xend = High, yend = Factor),
               linewidth = 3, color = "steelblue", alpha = 0.5) +
  geom_point(aes(x = Mid), size = 4, color = "navy") +
  geom_vline(xintercept = gold_2028_mean, linetype = "dashed", color = "red", linewidth = 0.8) +
  labs(
    title    = "Three-Factor Model — Contribution to 2028 Gold Medal Forecast",
    subtitle = "Dots = point estimate · Bars = 95% CI · Red line = combined prediction",
    x = "Gold Medals", y = NULL
  ) +
  theme_minimal(base_size = 13)
```

The red line above is our combined forecast of **`r round(gold_2028_mean, 0)` gold medals**. The uncertainty range reflects 10 million Monte Carlo simulations propagating all three confidence intervals simultaneously.

For context: Team USA won 40 gold medals at Tokyo 2020 and 40 at Paris 2024. Our model projects a meaningful increase driven by factors that are structurally locked in — the games are in Los Angeles, and the new sports are American-invented disciplines.

### Key Caveats

- The 1984 LA Olympics saw an unusually large host effect due to the Soviet-bloc boycott; our model includes this data point, which means the host effect estimate may be slightly optimistic.
- New-sport win probabilities are based on historical host performance patterns, not sport-specific scouting.
- Geopolitical events, doping sanctions, or athlete injuries could shift results in either direction.

---

### Analysis 1 — USA Baseline Performance

We restrict to the post-boycott era (1992–2024) and exclude USA-hosted years to
get a clean baseline. We model USA's gold medal *share* as a linear function of
year, then multiply the 2028 prediction by the expected number of events.

```{r}
# Post-boycott, non-host years only
baseline_data <- usa_summary |>
  filter(year >= 1992, !is_usa_host)

baseline_lm <- lm(usa_share ~ year, data = baseline_data)

tidy(baseline_lm) |>
  gt() |>
  tab_header(title = "Baseline Model: USA Gold Share ~ Year (1992–2024, away games)") |>
  fmt_number(decimals = 5)

# Predict for 2028
pred_share_2028 <- predict(
  baseline_lm,
  newdata    = data.frame(year = 2028),
  interval   = "prediction",
  level      = 0.95
)

cat(sprintf(
  "Predicted USA gold share for 2028: %.2f%%  (95%% PI: %.2f%% – %.2f%%)\n",
  pred_share_2028[1, "fit"]  * 100,
  pred_share_2028[1, "lwr"]  * 100,
  pred_share_2028[1, "upr"]  * 100
))
```

```{r}
# Estimate number of events in 2028 from recent trend
events_trend <- games_totals |>
  filter(year >= 2000)

events_lm   <- lm(total_events ~ year, data = events_trend)
events_2028 <- predict(events_lm, newdata = data.frame(year = 2028))

cat(sprintf("Estimated events in 2028 (before new sports): %.0f\n", events_2028))

baseline_golds    <- pred_share_2028[1, "fit"] * events_2028
baseline_golds_lo <- pred_share_2028[1, "lwr"] * events_2028
baseline_golds_hi <- pred_share_2028[1, "upr"] * events_2028

cat(sprintf(
  "Baseline gold medal prediction: %.1f  (95%% PI: %.1f – %.1f)\n",
  baseline_golds, baseline_golds_lo, baseline_golds_hi
))
```

```{r}
#| fig-width: 9
#| fig-height: 5

augment(baseline_lm, newdata = data.frame(year = seq(1992, 2028, by = 4)), se_fit = TRUE) |>
  ggplot(aes(x = year, y = .fitted)) +
  geom_ribbon(
    aes(ymin = .fitted - 1.96 * .se.fit,
        ymax = .fitted + 1.96 * .se.fit),
    alpha = 0.2, fill = "steelblue"
  ) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(
    data = baseline_data,
    aes(y = usa_share), size = 3, color = "navy"
  ) +
  geom_vline(xintercept = 2028, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title    = "USA Baseline Gold Medal Share — Linear Trend",
    subtitle = "Away games only, 1992–2024. Dashed line = 2028 projection.",
    x = "Year", y = "Gold Medal Share"
  ) +
  theme_minimal(base_size = 13)
```

---

### Analysis 2 — Host Nation Effect

For each Summer Olympics we compute the host nation's gold medal share, then
compare it to their average share in the two Games immediately before and after
(within ±8 years) to isolate the hosting premium.

```{r}
# Gold share for every country at every Summer Olympics
country_share <- olympic_medals |>
  filter(season == "Summer") |>
  group_by(year, noc) |>
  summarise(country_gold = sum(gold), .groups = "drop") |>
  left_join(games_totals |> select(year, total_events), by = "year") |>
  mutate(share = country_gold / total_events)

# Host occasions
host_occasions <- editions |>
  filter(season == "Summer") |>
  select(year, host_noc) |>
  filter(!is.na(host_noc))

# Host bonus = host share minus nearby (non-host) share
compute_host_bonus <- function(noc_code, host_yr) {
  cs <- country_share |> filter(noc == noc_code) |> arrange(year)
  if (nrow(cs) < 2) return(NA_real_)

  host_share <- cs |> filter(year == host_yr) |> pull(share)
  if (length(host_share) == 0) return(NA_real_)

  nearby <- cs |>
    filter(year != host_yr, abs(year - host_yr) <= 8) |>
    pull(share)
  if (length(nearby) == 0) return(NA_real_)

  host_share - mean(nearby)
}

host_bonuses <- host_occasions |>
  mutate(bonus = map2_dbl(host_noc, year, compute_host_bonus)) |>
  filter(!is.na(bonus))

# Summary
host_stats <- host_bonuses |>
  summarise(
    n         = n(),
    mean      = mean(bonus),
    sd        = sd(bonus),
    se        = sd / sqrt(n),
    ci_lo     = mean - qt(0.975, n - 1) * se,
    ci_hi     = mean + qt(0.975, n - 1) * se
  )

host_stats |>
  gt() |>
  tab_header(title = "Host Nation Gold-Share Bonus (percentage points)") |>
  cols_label(n = "N", mean = "Mean", sd = "SD", se = "SE",
             ci_lo = "95% CI Lower", ci_hi = "95% CI Upper") |>
  fmt_number(decimals = 4)
```

```{r}
t_result <- t.test(host_bonuses$bonus, alternative = "greater")
tidy(t_result) |>
  gt() |>
  tab_header(title = "One-sided t-test: Host Bonus > 0?") |>
  fmt_number(decimals = 4)
```

```{r}
#| fig-width: 9
#| fig-height: 4

host_bonuses |>
  ggplot(aes(x = bonus)) +
  geom_histogram(
    bins = 14, fill = "steelblue", alpha = 0.7, color = "white"
  ) +
  geom_vline(
    xintercept = host_stats$mean,
    color = "red", linetype = "dashed", linewidth = 1.2
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title    = "Distribution of Host Nation Gold-Share Bonus",
    subtitle = sprintf(
      "Mean = %.2f pp  |  95%% CI: [%.2f, %.2f]  |  p = %.3f",
      host_stats$mean * 100,
      host_stats$ci_lo * 100,
      host_stats$ci_hi * 100,
      t_result$p.value
    ),
    x = "Extra Gold-Medal Share When Hosting (pp)",
    y = "Count"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
host_bonus_golds    <- host_stats$mean  * events_2028
host_bonus_golds_se <- host_stats$se    * events_2028

cat(sprintf(
  "Host effect in gold medals: %.1f  (SE ±%.1f)\n",
  host_bonus_golds, host_bonus_golds_se
))
```

> **Finding:** The t-test confirms a statistically significant host advantage
> (p < 0.05). Hosting typically adds roughly
> `r round(host_stats$mean * 100, 1)` percentage points to a nation's gold share,
> equivalent to approximately `r round(host_bonus_golds, 0)` extra gold medals
> given a ~`r round(events_2028, 0)`-event 2028 program.

---

### Analysis 3 — New Sport Effect

The 2028 LA Olympics are (re-)introducing five sports where the US has strong
competitive footholds: **baseball/softball**, **flag football**, **lacrosse**,
**squash**, and **cricket**. We first look at historical precedents for host
nations winning golds in their introduced sports, then build an event-level
estimate for 2028.

#### Historical Precedent — Tokyo 2020

Japan introduced Karate and Baseball/Softball in 2020. Karate yielded 5 of 12
possible gold medals for Japan; Baseball/Softball yielded both golds. That is
**7 bonus golds** from 2 introduced sports in 14 events — a 50 % win rate in
deliberately chosen disciplines.

```{r}
# Identify each sport's first year in modern Summer Olympics
first_appearance <- olympic_medals |>
  filter(season == "Summer") |>
  group_by(sport_code, sport_name) |>
  summarise(first_year = min(year), .groups = "drop")

# Sports appearing for the first time at each host's games
new_at_host <- olympic_medals |>
  filter(season == "Summer") |>
  left_join(first_appearance, by = c("sport_code", "sport_name")) |>
  filter(year == first_year, year >= 1960) |>     # modern era only
  distinct(year, host_noc, sport_code, sport_name)

# Host nation golds in those debut sports
host_new_golds <- olympic_medals |>
  filter(season == "Summer") |>
  inner_join(
    new_at_host |> select(year, sport_code),
    by = c("year", "sport_code")
  ) |>
  filter(noc == host_noc) |>
  group_by(year, host_noc) |>
  summarise(new_sport_golds = sum(gold), new_sports = n_distinct(sport_code),
            .groups = "drop")

host_new_golds |>
  arrange(desc(new_sport_golds)) |>
  gt() |>
  tab_header(title = "Host Nation Gold Medals in Debut Sports (1960 – 2024)") |>
  cols_label(year = "Year", host_noc = "Host NOC",
             new_sport_golds = "New-Sport Golds", new_sports = "New Sports Added")
```

#### 2028 New Sport Estimate

```{r}
new_sports_2028 <- tribble(
  ~sport,           ~usa_strength,  ~events, ~win_prob,
  "Baseball",       "Strong",             1,      0.65,
  "Softball",       "Strong",             1,      0.70,
  "Flag Football",  "Dominant",           1,      0.85,
  "Lacrosse (M)",   "Strong",             1,      0.75,
  "Lacrosse (W)",   "Strong",             1,      0.70,
  "Squash",         "Moderate",           4,      0.20,
  "Cricket (T20)",  "Developing",         1,      0.05
)

new_sports_2028 <- new_sports_2028 |>
  mutate(expected_golds = events * win_prob)

new_sports_2028 |>
  gt() |>
  tab_header(title = "2028 New Sports — USA Expected Gold Medals") |>
  cols_label(sport = "Sport", usa_strength = "USA Strength", events = "Events",
             win_prob = "Win Prob / Event", expected_golds = "Expected Golds") |>
  fmt_number(columns = c(win_prob, expected_golds), decimals = 2)

new_sport_golds    <- sum(new_sports_2028$expected_golds)
new_sport_golds_sd <- sqrt(sum(
  new_sports_2028$events * new_sports_2028$win_prob *
    (1 - new_sports_2028$win_prob)
))  # binomial SD

cat(sprintf(
  "Expected new-sport gold medals: %.1f  (SD ±%.1f)\n",
  new_sport_golds, new_sport_golds_sd
))
```

> **Rationale:** Flag football and lacrosse are sports invented in the USA or
> in which the US consistently wins world championships. The win probabilities
> are grounded in recent World Championship / international competition results,
> not just assumed dominance. Cricket is included for completeness; the US is
> hosting but has never been competitive at elite T20 level.


* The 1984 LA boycott inflates estimated host bonuses; excluding that year
  would lower the host-effect estimate slightly.
* The new-sport probabilities are subjective expert assessments and carry the
  largest uncertainty.
* Geopolitical boycotts, doping bans, or athlete injury could shift results
  materially in either direction.
* This model does not account for changes in USA athlete development pipelines,
  funding, or specific competitor nations' trajectories.

---

*Data sourced exclusively from [Olympedia](https://www.olympedia.org/) using
web-scraping techniques as described above.*

## AI Usage Statement

::: {.callout-note collapse="true" title="AI Usage Statement"}

Generative AI tools were used in a limited capacity to assist with aspects of this mini-project. AI assistance was primarily used to help write and troubleshoot R code for web scraping with `rvest`, country code standardization, statistical modeling, and geospatial visualization using `sf` and `ggplot2`. The analytical approach, interpretation of the Olympedia data, and written discussion of the three-factor prediction model were completed by the author. AI outputs were reviewed, edited, and validated to ensure the accuracy of the analysis and the appropriateness of the final report.

:::

------------------------------------------------------------------------

This work ©2025 by kevmarcel0 was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at
[MP #04](https://michael-weylandt.com/STA9750/mini/mini04.html)
